Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-01-12 analysis/paper/article.qmd
BM responding to JB's comments

title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 900
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Many archaeology graduate students pursue advanced degrees in the hope of landing an academic career. Job listing websites often serve as the first place for these students when seeking academic positions. We examined tenure-track job advertisements over the past decade to gain insights into the academic job market for archaeologists. Using data from the community-edited Academic Jobs Wiki for Archaeology, we examine changes in the academic job market over time. We analyzed the text of 431 job ads posted from 2013--2023. Our analysis focuses on shifts in archaeological topics, methods and geographic regions specified in archaeological job ads. We also investigate whether the labor burden for applicants has changed over time: do institutions request more information and documents from applicants at the initial stages of application, compared to a decade ago? We assess the influence of socio-political factors on the changing focus of research topics in the field. This research aims to assist current and future archaeology students and graduates in better understanding the job market and the requirements of employers, thereby aiding them in effectively preparing for their applications for positions in archaeology.
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

@@

# explore over time
fig_prop_by_job_title_per_year <- 
jobdata %>%
  group_by(year_ad_posted) %>%
  count(job_title_simple) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot() +
  aes(year_ad_posted,
      n,
      group = job_title_simple,
      fill = job_title_simple) +
 # geom_line(linewidth = 2) +
  geom_col() +
 theme_minimal(base_size = base_size) +
  xlab("") +
  ylab("Number of\ntenure track job ads") +
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(nrow=2,
                               byrow=TRUE,
                               "Job title",
                                keywidth=0.1,
                 keyheight=0.1,
                 default.unit="inch")) +
    theme(legend.position = c(0.5, 0.85)) +
  guides(x = "none") +
  ylim(0,100)

@@

fig_prop_by_job_title_per_year_us <- 
jobdata_us_only %>%
  group_by(year_ad_posted) %>%
  count(job_title_simple) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot() +
  aes(year_ad_posted,
      n,
      group = job_title_simple,
      fill = job_title_simple) +
 # geom_line(linewidth = 2) +
  geom_col() +
 theme_minimal(base_size = base_size) +
  xlab("") +
  ylab("Number of\ntenure track job ads") +
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(nrow=2,
                               byrow=TRUE,
                               "Job title",
                                keywidth=0.1,
                 keyheight=0.1,
                 default.unit="inch")) +
    theme(legend.position = c(0.5, 0.85)) +
  guides(x = "none") +
  ylim(0,100)

@@

# count of TT jobs per year from our manual data collection,
# join with our total number of all jobs by scraping
count_of_tt_jobs_per_year_from_our_form <- 
jobdata %>% 
  group_by(url_to_data_source_e_g_paste_in_url_to_the_jobs_wiki_page) %>% 
  tally() %>% 
  right_join(total_number_of_jobs_per_year_tbl) %>% 
  rename(n_tt_jobs  = n,
         n_total_jobs = total_number_of_jobs_per_year) %>% 
  mutate(n_non_tt_jobs = n_total_jobs - n_tt_jobs,
         ratio_tt_2_ntt = n_tt_jobs / n_non_tt_jobs) %>% 
  mutate(year = str_extract(url_to_data_source_e_g_paste_in_url_to_the_jobs_wiki_page,  "[[0-9]]{4}-[[0-9]]{4}|2021-22")) %>%  
  mutate(year = ifelse(year =="2021-22",  "2021-2022", year))  %>% 
  # add line break for prettier plotting
  mutate(year_ad_posted_break = str_replace(year, "-", "-\n"))

# draw plot
fig_ratio_tt_2_ntt_jobs_per_year <- 
  ggplot(count_of_tt_jobs_per_year_from_our_form) +
  aes(year_ad_posted_break, 
      group = 1,
      ratio_tt_2_ntt) +
  #geom_line(linewidth = 2) +
  geom_col() +
  geom_hline(yintercept = 1,
             colour = "red") +
  annotate("text", 
           x = 6, 
           y = 1.4, 
           label = "1:1 ratio",
           colour = "white",
           size = 3) +
  labs(y = "Ratio of tenure-track\nto non-tenure track and other",
       x = "") +
  theme_minimal(base_size = base_size) +
  scale_x_discrete(name = "Academic Year") 
   #guides(x = "none")

@@

```{r}
#| label: fig-show-basic-plots
#| fig-cap: "A: total number of job ads posted to the Academic Jobs Wiki for Archaeology in each year, with coloured sections showing the proportion of jobs by title and rank. B: Ratio of tenure-track to non-tenure-track positions over time.The proportion of the bar above the red line reflects the number of non-tenure track positions available relative to tenure-track positions."
#JB: I have added the last sentence above because I have not seen this kind of figure in an archaeology paper, so I think we should explain it. Will reduce friction for our readers.Please confirm that I am actually correctly explaining how the figure works, however.
#JB: I would be eternally grateful if you would convert the hyphens between the timespans on the x-axis into en dashes 😬

@@

```{=html}
<!--Comments on Abstract and Keywords: 
1. Is “jobs listing website” the appropriate category here? I often hear these referred to as “jobs boards”—that might be a better fit?
2. Structurally, the second research domain is phrased as a question. For consistency, it would be good to provide a similar structure for the first domain, e.g. “Our analysis focuses on the XXX—have the topics, methods, or geographic regions specified in archaeological job ads changed over time?”
3. Suggested keywords: archaeology, hiring, academia, job ads-->
```

<!--Ben, please note that for now I am keeping track of my references. I have bolded them in text below; I think I will want to learn how to use Zotero and incorporate the citations using your arcane Lovecraftian ritual language (BibTex?), so that for our next paper I am capable of participating in the entire workflow with less supervision-->

Attend any workshop, conference, or panel that includes early career researchers and the conversation will be steered inexorably towards the academic job market: who is hiring, who has attained a tenure-track position, and who is out of luck this season. This obsessive focus on career trajectories is warranted, particularly within the discipline of anthropology. Between 1985 and 2014, the number of PhDs produced by American anthropology departments increased by 70% (**Speakman et al., 2018a**), but the number of tenure-track positions advertised has stagnated or declined*.* As of2019, 53.5% of colleges and universities had replaced tenure-eligible lines with contingent positions (**AAUP 2022**); today, 71% of faculty in the US are non-tenure-track(**Culver and Kezar 2022***).* The erosion of permanent jobs in American higher education is linked to a complex intersection of political and economic factors (**Beck, under review**), including decreasing federal support for higher education, concomitant increases in student debt (**Gusterson 2017**), a pronounced shift in university investment from faculty to administration (**Graeber 2018: 162–163**), and the growing privatization and market orientation of scientific research (**Mirowski 2011**). Whatever the cause, research within anthropology has consistently shown that at most 20% of PhDs in the discipline are obtaining jobs in American anthropology departments (**Speakman et al. 2018b**), while the top-ranking programmes are placing fewer than one in three graduates in tenure-track jobs (**Mackie and Rockwell 2023**).

These cultural and economic shifts are also well encapsulated by anecdata; any conversation with a junior scholar will reveal the impact of these larger professional transformations on personal and professional trajectories. The first author on this paper has applied to five academic positions over the course of his career, while the last author has applied to one hundred. A twenty-fold increase in the number of applications required to obtain a workable long-term academic position may seem preposterous, but these changing requirements for applicants are familiar to anyone who has obtained their PhDs since the Great Recession. This competitive landscape has deeper roots than the last economic crisis; writing more than three decades ago, Rabinow pinpointed “an awareness… that standards have changed during the last thirty years and the quantitative and qualitative demands for entry into the system are immeasurably higher now” (**1992: 66**).

These higher demands include increased overall productivity, higher mobility, sometimes requiring bothnational and international moves, more time spent in precarious short-term contracts, and the investment of a greater proportion of time and energy in applying for long-term positions. In their survey of early career researchers in European archaeology, for example, Brami et al. found that 52% of respondents who had obtained their PhD at least two years before the survey were on fixed-term contracts, while only 9% were in permanent positions (**2023: Figure 5**). In their words, “the queue of eager post-docs hoping for a long-term appointment is getting longer, as is the average time spent in the ‘queue’” (**2023: 239**). Ribeiro and Giamakis emphasize that this demographic comprises an academic precariat that is essential to the functioning of universities, but consists of a workforce on short-term contracts perpetually “stranded between employment and unemployment” (**2023:10**).

Early career scholars with faculty ambitions are thus stuck in a double-bind, forced to funnel their energy into seemingly endless applications for which they receive little to no feedback and which remain invisible on their CVs. As Dennis et al. underscore, early career researchers are “supplicants”: “There is no way for applicants to point out that the job ads are taking too much time out of the scholarly community’s collective time bank” (**2022: 1**). Indeed, acknowledging the new reality that applying for academic positions is in and of itself a full-time job, a growing number of anthropology graduate programmes are offering professionalization courses that cover the ins and outs of everything from crafting strong cover letters and CVs to performing well in Zoom interviews and teaching demonstrations. These courses often provide valuable training in the hidden curriculum of the academy, but that job applications are so complicated that they require training to navigate suggests more disciplinary attention should be paid to the “market” as a historically contingent set of cultural practices. As Rabinow maintains, “if we want ethical considerations to play a central role in the articulation of truth and power—and I think we do—then bringing such considerations into view is the necessary first step towards recognizing who we are today and setting out on the road to a better place" (**1992: 71**) **.**

@@

The academic job market is a source of uncertainty, for both job seekers and hiring committees. Job seekers must search for ads that are well-matched to the skill sets and professional experiences that they have taken years to cultivate. Hiring committeesmust sift through tens to hundreds of applicants in the hope of finding a candidate that meets all of their needs and helps to realize their visions of their program and the discipline. Finding the right candidate for the position—or the right position for the candidate—is akin to finding a needle in a haystack.

@@

PhD graduates who do not go directly into a tenure-track faculty position after graduation often go into a series of short-term appointments as part-time or limited contract instructors, also known as adjunct or contingent faculty. Rising numbers of contingent faculty has long been a concern in US higher education [@Trevithick2010]. Data from the American Association of University Professors' 2021–22 faculty survey indicate that more than 60% of faculty positions in US universities were held by non-tenure track full time or part time contingent faculty members [@colby2022annual]. Contingent faculty positions are precarious because they provide low or no health and retirement benefits, little or nothing in the way of professional development, and lower salaries relative to tenure-track positions. PhD graduates who have a sequence of short-term academic positions are often disadvantaged financially, due to low compensation combined with expensive frequent relocation, socially due to isolation from family and community, and professionally due to being ineligible for many decision-making roles at the universities they work at [@platzerAcademicPrecarityAmerican2018].

Success in tenure-track job applications in archaeology is strongly determined by where applicants get their PhDs. Data from the 2014–2015 AnthroGuide publication of the American Anthropological Association shows that just ten out of over 100 US graduate programs produced over 30% of the graduates hired into tenure-track faculty positions [@speakman2018choosing]. Similarly, a network analysis of 1,918 faculty holding tenured or tenure-track positions at PhD-granting anthropology programs in the US in 2015 (including 506 archaeologists) found that the just fifteen graduate programs produced 52.8 percent of tenured and tenure-track positions [@kawaSocialNetworkUS2019]. This network analysis showedthat programs with large endowments and with faculty who hold prestigious awards and are widely cited by other scholars produce the majority of tenured and tenure-track faculty in anthropology. Hiring bias predicated on the prestige of specific institutions and programs is not unique to anthropology. Targeted studies of the disciplines of sociology, communication, and operations research, and industrial systems engineering show similar trends  (**Barnett et al., 2010; Castillo et al., 2018; Feeley and Tutzauer 2022; Nevin 2019**). **Broader analyses comparing X, Y and Z, and analyses while a recent analysis of XXX showed Z (Clauset et al., 2015;** [@wapmanQuantifyingHierarchyDynamics2022a]). Within anthropology, however, such studies draw attention to the paradox of a discipline committed to fighting social inequalities continuing to reproduce its own systematic inequalities through hiring practices that favor an elite minority of graduates with prestigious institutional affiliations.

One way that some hiring committees are tackling these biases towards prestigious institutions is by providing detailed instructions to applicants on how to prepare their application materials. In theory, detailed applications will allow the hiring committee to focus on evaluating the accomplishments of candidates across common categories, rather than ranking based on prestige signals in a CV, such as a name of the applicant's graduate program. This push for equity has resulted in job ads that are often highly prescriptive in the types of documents that applicants should submit. For example, in addition to a cover letter and CV, job ads in many fields now require applicants to submit short statements detailing their previous and future contributions to teaching, research, and diversity within their discipline. In a comparison of job ads from 1999–2000 and 2019–2020 published in *Anthropology News* @HelloTristesTropes2021 noticed an increase in the number of materials requested from applicants. For example, twice as many 2019–2020 job ads requested writing samples compared to 1999–2000, and nearly four times as many requested statements of teaching philosophy. In their review of the 'worst job ads of 2021' @dennisWorstAnthroJob2022 reported two job ads with particularly extreme requirements<!--Are you opposed to naming and shaming these institutions, which were Oberlin College and Grinnell College? I applied for one of these positions, and have no issue referring to them by name...-->, which respectively requested nine and thirteen documents from applicants. This high burden on applicants disproportionately favors people with more time and financial resources to prepare the required materials.

@@

**While attention to equity in job market materials is commendable, work on hiring practices suggests ZZZ. For example, Warren 2019 has shown XXX. Rivera, in a study of hiring practices at YYY, has likewise demonstrated ZZZ. Somewhere in this section I want to discuss the results from sociology, specifically the Warren 2019 piece and the Rivera 2017 piece on hiring practices.**

@@

Our primary data source is the Archaeology Academic Jobs Wiki<!--Check if Kelsky has anything too say about this. I also think there have been some articles about the pros and cons of the poli sci or Econ versions of this.-->. Originating in 2007, this is a set of freely accessible web pages that anyone can edit (anonymously or with a free user account) hosted by Fandom, a for-profit company. The Archaeology pages are part of the Academic Jobs Wiki, which coordinates similar collaboratively-edited resources for around 40 academic disciplines. The coordinators and contributors are nearly all anonymous or pseudonymous. Typically, contributors copy and paste the text of job ads from other sources, such as the *Chronicle of Higher Education*, *Higher Ed Jobs*, and university websites, into the wiki, collecting <!--consolidating? centralizing?--> ads originally posted in numerous different locations. Other contributors then edit the web page to add comments below an ad to share relevant information based on their experience in applying for that position, such as a tally of how many people have applied, the dates of events such as requests for more materials, interviews, offer made, rejection notices, etc. Contributors also edit the page to ask and answer questions about the positions and the application process. These comments make the Academic Jobs Wiki a unique resource for timely and specific information for job-seekers about positions they are interested in, and one of the most important internet resources for the academic job market. Because of its reputation for aggregating ads from diverse sources and rapidly-updated information that is not available elsewhere, the Academic Jobs Wiki has a large community of users that keep it updated and accurate has become an authoritative data source for studies of hiring trends in academia [e.g. @Passalacqua_2018; @musial2018five] and a widely recommended resource [e.g. @lightfoot2021preparing].

For each tenure-track job advertised on the Archaeology Academic Jobs Wiki during 2013-2023, we read the text and recorded the name of the hiring institution, the title of the position, and exact words and phrases from the ad about the topical, geographic, and methods foci on the position into a Google form. The topical focus is what we understood as the primary intellectual focus of the position—**examples of topical foci included X,Y, and Z**. The geographic focus is the region of the world that the ideal candidate has scholarly expertise on, **for example, X, Y, or Z**. The methods focus is the data-generating sub-field of archaeology that is mentioned in the ad. **Examples of methods used in this study include X, Y, or Z.** <!--It might be helpful to make this into a summary table with examples of all foci generated by the study, and then one showing the simplified 10–15 topics. I understand this is in the supplementary tables already.--> We also recorded the type and number of documents requested in each ad (e.g. cover letter, CV, statements on research, teaching, diversity<!-- Were these all relatively easy to recognise? I’ve applied to places like Penn State, for example, that do a stupid combined research/teaching/diversity statement, just to make applicants do extra work, seemingly.-->, syllabi, course descriptions, writing samples, transcripts) and how many names/letters of recommenders were requested in the ad.

>>>>

title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 900
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Many archaeology graduate students pursue advanced degrees in the hope of landing an academic career. Job listing websites often serve as the first place for these students when seeking academic positions. We examined tenure-track job advertisements over the past decade to gain insights into the academic job market for archaeologists. Using data from the community-edited Academic Jobs Wiki for Archaeology, we examine changes in the academic job market over time. We analyzed the text of 431 job ads posted from 2013--2023. Our analysis addresses the question of how have archaeological topics, methods, and geographic regions specified in archaeological job ads shifted over time? We also investigate whether the labor burden for applicants has changed over time: do institutions request more information and documents from applicants at the initial stages of application, compared to a decade ago? We assess the influence of socio-political factors on the changing focus of research topics in the field. This research aims to assist current and future archaeology students and graduates in better understanding the job market and the requirements of employers, thereby aiding them in effectively preparing for their applications for positions in archaeology.
keywords: |
  archaeology; hiring; academia; job ads 
highlights: |
  These are the highlights. 
---

```{=html}
<!--Comments on Abstract and Keywords: 
1. Is “jobs listing website” the appropriate category here? I often hear these referred to as “jobs boards”—that might be a better fit? BM: I agree that 'job board' is in common use, but I prefer 'jobs listing website' because it's more precise and explicit about the medium and contents

2. Structurally, the second research domain is phrased as a question. For consistency, it would be good to provide a similar structure for the first domain, e.g. “Our analysis focuses on the XXX—have the topics, methods, or geographic regions specified in archaeological job ads changed over time?” BM: agreed, I updated the abstract

3. Suggested keywords: archaeology, hiring, academia, job ads BM: added, you can also freely add these kinds of things directly, no worries -->
```

@@

# explore over time
fig_prop_by_job_title_per_year <- 
jobdata %>%
  group_by(year_ad_posted) %>%
  count(job_title_simple) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot() +
  aes(year_ad_posted,
      n,
      group = job_title_simple,
      fill = job_title_simple) +
 # geom_line(linewidth = 2) +
  geom_col() +
 theme_minimal(base_size = base_size) +
  xlab("") +
  ylab("Number of\ntenure-track job ads") +
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(nrow=2,
                               byrow=TRUE,
                               "Job title",
                                keywidth=0.1,
                 keyheight=0.1,
                 default.unit="inch")) +
    theme(legend.position = c(0.5, 0.85)) +
  guides(x = "none") +
  ylim(0,100)

@@

fig_prop_by_job_title_per_year_us <- 
jobdata_us_only %>%
  group_by(year_ad_posted) %>%
  count(job_title_simple) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot() +
  aes(year_ad_posted,
      n,
      group = job_title_simple,
      fill = job_title_simple) +
 # geom_line(linewidth = 2) +
  geom_col() +
 theme_minimal(base_size = base_size) +
  xlab("") +
  ylab("Number of\ntenure-track job ads") +
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(nrow=2,
                               byrow=TRUE,
                               "Job title",
                                keywidth=0.1,
                 keyheight=0.1,
                 default.unit="inch")) +
    theme(legend.position = c(0.5, 0.85)) +
  guides(x = "none") +
  ylim(0,100)

@@

# count of TT jobs per year from our manual data collection,
# join with our total number of all jobs by scraping
count_of_tt_jobs_per_year_from_our_form <- 
jobdata %>% 
  group_by(url_to_data_source_e_g_paste_in_url_to_the_jobs_wiki_page) %>% 
  tally() %>% 
  right_join(total_number_of_jobs_per_year_tbl) %>% 
  rename(n_tt_jobs  = n,
         n_total_jobs = total_number_of_jobs_per_year) %>% 
  mutate(n_non_tt_jobs = n_total_jobs - n_tt_jobs,
         ratio_tt_2_ntt = n_tt_jobs / n_non_tt_jobs) %>% 
  mutate(year = str_extract(url_to_data_source_e_g_paste_in_url_to_the_jobs_wiki_page,  "[[0-9]]{4}-[[0-9]]{4}|2021-22")) %>%  
  mutate(year = ifelse(year =="2021-22",  "2021-2022", year))  %>% 
  # add line break for prettier plotting
  mutate(year_ad_posted_break = str_replace(year, "-", 
                                            "–\n")) # en-dash

# draw plot
fig_ratio_tt_2_ntt_jobs_per_year <- 
  ggplot(count_of_tt_jobs_per_year_from_our_form) +
  aes(year_ad_posted_break, 
      group = 1,
      ratio_tt_2_ntt) +
  #geom_line(linewidth = 2) +
  geom_col() +
  geom_hline(yintercept = 1,
             colour = "red") +
  annotate("text", 
           x = 6, 
           y = 1.4, 
           label = "1:1 ratio",
           colour = "white",
           size = 3) +
  labs(y = "Ratio of tenure-track\nto non-tenure-track and other",
       x = "") +
  theme_minimal(base_size = base_size) +
  scale_x_discrete(name = "Academic Year") 
   #guides(x = "none")
```

@@

```{r}
#| label: fig-show-basic-plots
#| fig-cap: "A: total number of job ads posted to the Academic Jobs Wiki for Archaeology in each year, with coloured sections showing the proportion of jobs by title and rank. B: Ratio of tenure-track to non-tenure-track positions over time. The red line indicates a 1:1 ratio of tenure-track to non-tenure-track positions, bars taller than that line indicate more tenure-track than non-tenure-track positions in that year"
#JB: I have added the last sentence above because I have not seen this kind of figure in an archaeology paper, so I think we should explain it. Will reduce friction for our readers.Please confirm that I am actually correctly explaining how the figure works, however. BM: I've edited the caption further to clarify how to read that red line. 
#JB: I would be eternally grateful if you would convert the hyphens between the timespans on the x-axis into en dashes 😬 BM: done, you may commence eternal gratitude :)

@@

<!--Ben, please note that for now I am keeping track of my references. I have bolded them in text below; I think I will want to learn how to use Zotero and incorporate the citations using your arcane Lovecraftian ritual language (BibTex?), so that for our next paper I am capable of participating in the entire workflow with less supervision. BM: ok great! Yes, BibTeX, I believe you will love it. -->

Attend any workshop, conference, or panel that includes early career researchers and the conversation will be steered inexorably towards the academic job market: who is hiring, who has attained a tenure-track position, and who is out of luck this season. This obsessive focus on career trajectories is warranted, particularly within the discipline of anthropology. As of 2019, 53.5% of colleges and universities had replaced tenure-eligible lines with contingent positions (**AAUP 2022**); today, 71% of faculty in the US are non-tenure-track(**Culver and Kezar 2022***).* The erosion of permanent jobs in American higher education is linked to a complex intersection of political and economic factors (**Beck, under review**), including decreasing federal support for higher education, concomitant increases in student debt (**Gusterson 2017**), a pronounced shift in university investment from faculty to administration (**Graeber 2018: 162–163**), and the growing privatization and market orientation of scientific research (**Mirowski 2011**).

These cultural and economic shifts are also well encapsulated by anecdata; any conversation with a junior scholar will reveal the impact of these larger professional transformations on personal and professional trajectories. The first author on this paper has applied to five academic positions over the course of his career, while the last author has applied to one hundred. A twenty-fold increase in the number of applications required to obtain a workable long-term academic position may seem preposterous, but these changing requirements for applicants are familiar to anyone who has obtained their PhDs since the Great Recession of 2008. This competitive landscape has deeper roots than the last economic crisis; writing more than three decades ago, Rabinow pinpointed "an awareness... that standards have changed during the last thirty years and the quantitative and qualitative demands for entry into the system are immeasurably higher now" (**1992: 66**).

These higher demands include increased overall productivity, higher mobility, sometimes requiring bothnational and international moves, more time spent in precarious short-term contracts, and the investment of a greater proportion of time and energy in applying for long-term positions. In their survey of early career researchers in European archaeology, for example, Brami et al. found that 52% of respondents who had obtained their PhD at least two years before the survey were on fixed-term contracts, while only 9% were in permanent positions (**2023: Figure 5**). In their words, "the queue of eager post-docs hoping for a long-term appointment is getting longer, as is the average time spent in the 'queue'" (**2023: 239**). Ribeiro and Giamakis emphasize that this demographic comprises an academic precariat that is essential to the functioning of universities, but consists of a workforce on short-term contracts perpetually "stranded between employment and unemployment" (**2023:10**).

Early career scholars with faculty ambitions are thus stuck in a double-bind, forced to funnel their energy into seemingly endless applications for which they receive little to no feedback and which remain invisible on their CVs. As Dennis et al. underscore, early career researchers are "supplicants": "There is no way for applicants to point out that the job ads are taking too much time out of the scholarly community’s collective time bank" (**2022: 1**). Indeed, acknowledging the new reality that applying for academic positions is in and of itself a full-time job, a growing number of anthropology graduate programmes are offering professionalization courses that cover the ins and outs of everything from crafting strong cover letters and CVs to performing well in online video interviews and teaching demonstrations. These courses often provide valuable training in the hidden curriculum of the academy, but that job applications are so complicated that they require training to navigate suggests more disciplinary attention should be paid to the "market" as a historically contingent set of cultural practices. As Rabinow maintains, "if we want ethical considerations to play a central role in the articulation of truth and power—and I think we do—then bringing such considerations into view is the necessary first step towards recognizing who we are today and setting out on the road to a better place" (**1992: 71**) **.**

@@

<!-- BM: I think we have a little repetition in the Intro and Background sections, particulary in the mentions of data from papers by Speakman, Kawa et al. and Mackie & Rockwell. To reduce duplication, I've removed most of that from the intro, and  kept/shifted it into the background so we have a logical progression of general to more specific observations about the job market -->

The academic job market is a source of uncertainty, for both job seekers and hiring committees. Job seekers must search for ads that are well-matched to the skill sets and professional experiences that they have taken years to cultivate. Hiring committees must sift through tens to hundreds of applicants in the hope of finding a candidate that meets all of their needs and helps to realize their visions of their program and the discipline. Finding the right candidate for the position—or the right position for the candidate—is akin to finding a needle in a haystack.

@@

PhD graduates who do not go directly into a tenure-track faculty position after graduation often go into a series of short-term appointments as part-time or limited contract instructors, also known as adjunct or contingent faculty. Rising numbers of contingent faculty has long been a concern in US higher education [@Trevithick2010]. Data from the American Association of University Professors' 2021–22 faculty survey indicate that more than 60% of faculty positions in US universities were held by non-tenure-track full time or part time contingent faculty members [@colby2022annual]. Contingent faculty positions are precarious because they provide low or no health and retirement benefits, little or nothing in the way of professional development, and lower salaries relative to tenure-track positions. PhD graduates who have a sequence of short-term academic positions are often disadvantaged financially, due to low compensation combined with expensive frequent relocation, socially due to isolation from family and community, and professionally due to being ineligible for many decision-making roles at the universities they work at [@platzerAcademicPrecarityAmerican2018].

Success in tenure-track job applications in archaeology is strongly determined by where applicants get their PhDs. Data from the 2014–2015 AnthroGuide publication of the American Anthropological Association shows that just ten out of over 100 US graduate programs produced over 30% of the graduates hired into tenure-track faculty positions [@speakman2018choosing]. Top-ranking programmes are placing fewer than one in three graduates in tenure-track jobs [@mackieMarketShareAccounting2023]. Similarly, a network analysis of 1,918 faculty holding tenured or tenure-track positions at PhD-granting anthropology programs in the US in 2015 (including 506 archaeologists) found that the just fifteen graduate programs produced 52.8 percent of tenured and tenure-track positions [@kawaSocialNetworkUS2019]. This network analysis showed that programs with large endowments and with faculty who hold prestigious awards and are widely cited by other scholars produce the majority of tenured and tenure-track faculty in anthropology. Hiring bias predicated on the prestige of specific institutions and programs is not unique to anthropology. Targeted studies of the disciplines of sociology, communication, and operations research, and industrial systems engineering show similar trends (**Barnett et al., 2010; Castillo et al., 2018; Feeley and Tutzauer 2022; Nevin 2019**). **Broader analyses comparing X, Y and Z, and analyses while a recent analysis of XXX showed Z (Clauset et al., 2015;** [@wapmanQuantifyingHierarchyDynamics2022a]). Within anthropology, however, such studies draw attention to the paradox of a discipline committed to fighting social inequalities continuing to reproduce its own systematic inequalities through hiring practices that favor an elite minority of graduates with prestigious institutional affiliations.

One way that some hiring committees are tackling these biases towards prestigious institutions is by providing detailed instructions to applicants on how to prepare their application materials. In theory, detailed applications will allow the hiring committee to focus on evaluating the accomplishments of candidates across common categories, rather than ranking based on prestige signals in a CV, such as a name of the applicant's graduate program. This push for equity has resulted in job ads that are often highly prescriptive in the types of documents that applicants should submit. For example, in addition to a cover letter and CV, job ads in many fields now require applicants to submit short statements detailing their previous and future contributions to teaching, research, and diversity within their discipline. In a comparison of job ads from 1999–2000 and 2019–2020 published in *Anthropology News* @HelloTristesTropes2021 noticed an increase in the number of materials requested from applicants. For example, twice as many 2019–2020 job ads requested writing samples compared to 1999–2000, and nearly four times as many requested statements of teaching philosophy. In their review of the 'worst job ads of 2021' @dennisWorstAnthroJob2022 reported job ads from Oberlin College and Grinnell College with particularly extreme requirements <!--Are you opposed to naming and shaming these institutions, which were Oberlin College and Grinnell College? I applied for one of these positions, and have no issue referring to them by name... BM: nope, I've added their names -->, which respectively requested nine and thirteen documents from applicants. This high burden on applicants disproportionately favors people with more time and financial resources to prepare the required materials.

@@

**While attention to equity in job market materials is commendable, work on hiring practices suggests ZZZ. For example, Warren 2019 has shown XXX. Rivera, in a study of hiring practices at YYY, has likewise demonstrated ZZZ. Somewhere in this section I want to discuss the results from sociology, specifically the Warren 2019 piece and the Rivera 2017 piece on hiring practices.** <!-- ok, sounds good! -->

@@

Our primary data source is the Archaeology Academic Jobs Wiki<!--Check if Kelsky has anything too say about this. I also think there have been some articles about the pros and cons of the poli sci or Econ versions of this. BM: I did have a search of her blog where there are ~20 mentions, all pretty neutral, her most interesting comment is to notes that it originated in the year of the Great Recession -->. Originating in 2007, this is a set of freely accessible web pages that anyone can edit (anonymously or with a free user account) hosted by Fandom, a for-profit company. The Archaeology pages are part of the Academic Jobs Wiki, which coordinates similar collaboratively-edited resources for around 40 academic disciplines. The coordinators and contributors are nearly all anonymous or pseudonymous. Typically, contributors copy and paste all or some of the text of job ads from other sources, such as the *Chronicle of Higher Education*, *Higher Ed Jobs*, and university websites, into the wiki, collecting <!--consolidating? centralizing? BM: I prefer the shortest and simplest word that will do the job --> ads originally posted in numerous different locations. Other contributors then edit the web page to add comments below an ad to share relevant information based on their experience in applying for that position. These edits result in annotations such as a tally of how many people have applied, the dates of events such as requests for more materials, interviews, offer made, rejection notices, etc. Contributors also edit the page to ask and answer questions about the positions and the application process. These comments make the Academic Jobs Wiki a unique resource for timely and specific information for job-seekers about positions they are interested in, and one of the most important internet resources for the academic job market. Because of its reputation for aggregating ads from diverse sources and rapidly-updated information that is not available elsewhere, the Academic Jobs Wiki has a large community of users that keep it updated and accurate has become an authoritative data source for studies of hiring trends in academia [e.g. @Passalacqua_2018; @musial2018five] and a widely recommended resource [e.g. @lightfoot2021preparing].

For each tenure-track job advertised on the Archaeology Academic Jobs Wiki during 2013-2023, we read the text and recorded the name of the hiring institution, the title of the position, and exact words and phrases from the ad about the topical, geographic, and methods foci on the position into a Google form. The topical focus is what we understood as the primary intellectual focus of the position—examples of topical foci included environmental archaeology, public archaeology, and North American archaeology. The geographic focus is the region of the world that the ideal candidate has scholarly expertise on, for example, Southwest US, Mediterranean, or Asia and India. The methods focus is the data-generating sub-field of archaeology that is mentioned in the ad. Examples of methods used in this study include archaeobotany, lithic analysis, or zooarchaeology. <!--It might be helpful to make this into a summary table with examples of all foci generated by the study, and then one showing the simplified 10–15 topics. I understand this is in the supplementary tables already. BM: I've added in the 'for example' you requested, but feels very repetitive to give examples in this paragraph, and then the full set of topics etc. in the next paragraph, and add then add a table as well. It's like saying "we found 5000 lithics, one random one was 45 mm long, another random piece was 12 mm wide, etc." readers don't expect that in the manuscript  --> We also recorded the type and number of documents requested in each ad (e.g. cover letter, CV, statements on research, teaching, diversity <!-- Were these all relatively easy to recognise? I’ve applied to places like Penn State, for example, that do a stupid combined research/teaching/diversity statement, just to make applicants do extra work, seemingly. BM: we treated these kinds of combo statements as separate requests, I don't know how many there were, I'm ok to omit that detail -->, syllabi, course descriptions, writing samples, transcripts) and how many names/letters of recommenders were requested in the ad, if any.
